---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we load the data using read.csv(). It has 17,568 observations of 3 variables: steps, date, and interval.

```{r,echo=TRUE}
activity <- read.csv("activity/activity.csv")
```

Now that we have the data, we can process/transform it into a format suitable for analysis. To do this, I'd like to do is turn the dates into Date objects, and the intervals into factors.

```{r,echo=TRUE}
activity[,2] <- as.Date(activity[,2], format="%Y-%m-%d")
activity[,3] <- as.factor(activity[,3])
```

## What is mean total number of steps taken per day?

To answer this question, first I will find the total number of steps taken per day.

```{r,echo=TRUE}
stepsPerDay <- tapply(activity[,1], activity[,2], sum, na.rm = TRUE)
stepsPerDay

```

Then I will make a histogram of the total number of steps taken per day.

```{r,echo=TRUE}
hist(stepsPerDay)
```

Finally, I will calculate the mean and median of the total number of steps taken per day.

```{r,echo=TRUE}
mean(stepsPerDay)
median(stepsPerDay)
```

## What is the average daily activity pattern?

Let's make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r,echo=TRUE}
avgStepsPerInterval <- tapply(activity[,1], activity[,3], mean, na.rm = TRUE)
plot(avgStepsPerInterval, type = "l", xlab="5-minute interval", ylab="Average Number of Steps", main="Average Steps per 5-minute Interval")

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r,echo=TRUE}
maxSteps <- max(avgStepsPerInterval)
which(avgStepsPerInterval == maxSteps)
```

So, the 104th time interval (that is not entirely NA) which corresponds to the interval 835-840 minutes in the day.

## Imputing missing values

To understand the missing data, we can calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r,echo=TRUE}
sum(is.na(activity[,1]))
```

Now to devise a strategy for filling in all of the missing values in the dataset. I think I'll use average number of steps for that day, truncated to an integer.

```{r,echo=TRUE}
fillNA <- function(day){
    daySteps <- activity[activity$date==day,1]
    dayMean <- as.integer(mean(daySteps, na.rm = TRUE))
    if(is.na(dayMean)){
        dayMean <- 0
    }
    return(dayMean)
}
```

Create a new dataset that is equal to the original dataset but with the missing data filled in. I'll sum the NAs again to show you that they have all been replaced.

```{r,echo=TRUE}
for(i in seq_along(activity[,1])){
    if(is.na(activity[i,1])){
        activity[i,1] <- fillNA(activity[i,2])
    }
}
sum(is.na(activity[,1]))
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r,echo=TRUE}
stepsPerDay <- tapply(activity[,1], activity[,2], sum, na.rm = TRUE)
hist(stepsPerDay)
mean(stepsPerDay)
median(stepsPerDay)

```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "Weekday" and "Weekend" indicating whether a given date is a weekday or weekend day.

```{r,echo=TRUE}
activity$fun <- weekdays(activity$date)
for(i in seq_along(activity$fun)){
    if(activity[i,4] == "Saturday" || activity[i,4] == "Sunday"){
        activity[i,4] <- "Weekend"
    }
    else{
        activity[i,4] <- "Weekday"
    }
}
activity$fun <- as.factor(activity$fun)
summary(activity)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r,echo=TRUE}
weekendActivity <- activity[activity$fun == "Weekend",1:3]
weekdayActivity <- activity[activity$fun == "Weekday",1:3]

avgWeekendStepsPerInterval <- tapply(weekendActivity[,1], weekendActivity[,3], mean)
avgWeekdayStepsPerInterval <- tapply(weekdayActivity[,1], weekdayActivity[,3], mean)

rng <- range(avgWeekendStepsPerInterval, avgWeekdayStepsPerInterval)
par(mfrow = c(1,2))

plot(avgWeekendStepsPerInterval, type = "l", ylim = rng, xlab="5-minute interval", ylab="Average Number of Steps", main="Weekend")
plot(avgWeekdayStepsPerInterval, type = "l", ylim = rng, xlab="5-minute interval", ylab="Average Number of Steps", main="Weekday")
```
